Enforced Invariant Confluence

### 2018-04-01 doc/paper/sections/invariant_confluence.tex
Work on segmented invariant-confluence.

A \defword{distributed object} $O = (S, \join)$ consists of a set $S$ of
states and a binary merge operator $\join: S \times S \to S$ which merges two
states into one. A \defword{transaction} $t: S \to S$ is a function which maps
one state to another. An \defword{invariant} $I$ is a subset of $S$.
Notationally, we write $I(s)$ to denote that $s$ satisfies the invariant (i.e.
$s \in I$) and $\lnot I(s)$ to denote that $s$ does not satisfy the invariant
(i.e. $s \notin I$).
%
For example, $O = (\ints, +)$ is a distributed object consisting of integers
merged by addition; $t(x) = x + 1$ is a transaction which adds one to a state;
and $\setst{x \in \ints}{x \geq 0}$ is the invariant that states $x$ are
non-negative.

In our system model, a distributed object $O$ is replicated across a set $p_1,
\ldots, p_n$ of $n$ servers. Each server $p_i$ manages a replica $s_i \in S$ of
the replicated object. Every server begins with a start state $s_0 \in S$, a
fixed set $T$ of transactions, and an invariant $I$. Servers can perform one of
two actions.

@@

To define invariant-confluence formally, we represent a state produced by a
system execution as a simple expression generated by the grammar
\[
  e ::= s \mid t(e) \mid e_1 \join e_2
\]
where $s$ represents a state in $S$ and $t$ represents a transaction in $T$. As
an example, consider the system execution in \figref{SystemExecution} in which
a distributed object is replicated across servers $p_1$, $p_2$, and $p_3$.
Server $p_3$ begins with state $s_0$, transitions to state $s_2$ by executing
transaction $u$, transitions to state $s_5$ by executing transaction $w$, and
then transitions to state $s_7$ by merging with server $p_1$. Similarly, server
$p_1$ ends up with state $s_6$ after executing transactions $t$ and $v$ and
merging with server $p_2$. In \figref{Expression}, we see the abstract syntax
tree of the corresponding expression for state $s_7$.

>>>>

\subsection{System Model}
A \defword{distributed object} $O = (S, \join)$ consists of a set $S$ of
states and a binary merge operator $\join: S \times S \to S$ which merges two
states into one. A \defword{transaction} $t: S \to S$ is a function which maps
one state to another. An \defword{invariant} $I$ is a subset of $S$.
Notationally, we write $I(s)$ to denote that $s$ satisfies the invariant (i.e.
$s \in I$) and $\lnot I(s)$ to denote that $s$ does not satisfy the invariant
(i.e. $s \notin I$).
%
For example, $O = (\ints, +)$ is a distributed object consisting of integers
merged by addition; $t(x) = x + 1$ is a transaction which adds one to a state;
and $\setst{x \in \ints}{x \geq 0}$ is the invariant that states $x$ are
non-negative.

In our system model, a distributed object $O$ is replicated across a set $p_1,
\ldots, p_n$ of $n$ servers. Each server $p_i$ manages a replica $s_i \in S$ of
the replicated object. Every server begins with a start state $s_0 \in S$, a
fixed set $T$ of transactions, and an invariant $I$. Servers repeatedly perform
one of two actions.

@@

\subsection{Expression-Based Formalism}
To define invariant-confluence formally, we represent a state produced by a
system execution as a simple expression generated by the grammar
\[
  e ::= s \mid t(e) \mid e_1 \join e_2
\]
where $s$ represents a state in $S$ and $t$ represents a transaction in $T$. As
an example, consider the system execution in \figref{SystemExecution} in which
a distributed object is replicated across servers $p_1$, $p_2$, and $p_3$.
Server $p_3$ begins with state $s_0$, transitions to state $s_2$ by executing
transaction $u$, transitions to state $s_5$ by executing transaction $w$, and
then transitions to state $s_7$ by merging with server $p_1$. Similarly, server
$p_1$ ends up with state $s_6$ after executing transactions $t$ and $v$ and
merging with server $p_2$. In \figref{Expression}, we see the abstract syntax
tree of the corresponding expression for state $s_7$.

### 2018-04-01 doc/paper/sections/segmented_invariant_confluence.tex
Work on segmented invariant-confluence.

\section{Segmented Invariant-Confluence}\seclabel{SegmentedInvariantConfluence}

>>>>

\section{Segmented Invariant-Confluence}\seclabel{SegmentedInvariantConfluence}
If a distributed object is invariant-confluent, then the object can be
replicated without the need for any form of coordination to maintain the
object's invariant. But what if the object is \emph{not} invariant-confluent?
In this section, we present a generalization of invariant-confluence called
\defword{segmented invariant-confluence} that can be used to maintain the
invariants of non-invariant-confluent objects, requiring only a small amount of
coordination.

The main idea behind segmented invariant-confluence is to segment the state
space into a number of segments and restrict the set of allowable transactions
within each segment in such a way that the object is invariant-confluent
\emph{within each segment} (even though it may not be globally
invariant-confluent). Then, servers can run coordination-free within a segment
and need only coordinate when transitioning from one segment to another. We now
formalize this idea by defining segmented invariant-confluence and a new system
model to execute segmented invariant-confluent objects.

\subsection{Formalism}
Consider a distributed object $O = (S, \join)$, a start state $s_0 \in S$, a
set of transitions $T$, and an invariant $I$. A segmentation $\Sigma = (I_1,
T_1), \ldots, (I_n, T_n)$ is a sequence (not a set) of $n$ segments$(I_i, T_i)$
where every $T_i$ is a subset of $T$. $O$ is \defword{segmented
invariant-confluent} with respect to $s_0$, $T$, $I$, and $\Sigma$, abbreviated
\defword{\sTISconfluent}, if the following conditions hold:
\begin{itemize}
  \item
    The start state satisfies the invariant (i.e. $I(s_0)$).

  \item
    $I$ is covered by the invariants in $\Sigma$ (i.e. $I = \cup_{i=1}^n I_i$).
    Note that invariants in $\Sigma$ do \emph{not} have to be disjoint. Thus,
    they do not have to partition $I$; they just have to cover $I$.

  \item
    For every $(I_i, T_i) \in \Sigma$ and for every state $s \in I_i$, $O$ is
    \sticonfluent{s}{T_i}{I_i}.
\end{itemize}

{\input{figures/segmented_z2}}

\begin{example}\examplelabel{SegmentedZ2}
  Consider again the object $O = (\ints \times \ints, \join)$, transactions $T
  = \set{t_{x+1}, t_{y-1}}$, and invariant $I = \setst{(x, y)}{xy \leq 0}$ from
  \exampleref{Z2}, but now let the start state $s_0$ be $(-42, 42)$. $O$ is
  \emph{not} \sTISconfluent{} because the points $(0, 42)$ and $(42, 0)$ are
  reachable, and merging these points yields $(42, 42)$ which violates the
  invariant. However, $O$ is \sTISconfluent{} for $\Sigma = (I_1, T_1)$, $(I_2,
  T_2)$, $(I_3, T_3)$, $(I_4, T_4)$ where
  \begin{align*}
    I_1 &= \setst{(x, y)}{x < 0, y > 0}       &T_1 &= \set{t_{x+1}, t_{y-1}} \\
    I_2 &= \setst{(x, y)}{x \geq 0, y \leq 0} &T_2 &= \set{t_{x+1}, t_{y-1}} \\
    I_3 &= \setst{(x, y)}{x = 0}              &T_3 &= \set{t_{y-1}} \\
    I_4 &= \setst{(x, y)}{y = 0}              &T_4 &= \set{t_{x+1}}
  \end{align*}
  $\Sigma$ is illustrated in \figref{SegmentedZ2}. Clearly, $s_0$ satisfies the
  invariant, and $I_1, I_2, I_3, I_4$ cover $I$. Moreover, for every $(I_i,
  T_i) \in \Sigma$, we see that $O$ is \iclosed{I_i}, so $O$ is
  \sticonfluent{s}{T_i}{I_I} for every $s \in I_i$. Thus, $O$ is
  \sTISconfluent{}.
\end{example}

\subsection{System Model}
Now, we describe a new system model used to replicate a segmented
invariant-confluent object with coordination-free execution within a segment
and a small amount of coordination to transition between segments. As before,
we replicate an object $O$ across a set $p_1, \ldots, p_n$ of $n$ servers each
of which manages a replica $s_i \in S$ of the object. Every server begins with
$s_0$, $T$, $I$, and $\Sigma$. Moreover, at any given point in time, a server
designates one of the segments in $\Sigma$ as its \defword{active segment}.
Initially, every server chooses the first segment $(I_i, T_i) \in \Sigma$ such
that $I_i(s_0)$ to be its active segment. We'll see momentarily the
significance of the active segment.

As before, servers repeatedly perform one of two actions: execute a transaction
or merge states with another server. Merging states in the new system model is
identical to merging states in the old system model. A server $p_i$ sends its
state $s_i$ to another server $p_j$ which \emph{must} merge $s_i$ into its
state $s_j$.
%
Transaction execution in the new system model, on the other hand, is a bit more
involved. Consider a server $s_i$ with active segment $(I_i, T_i)$. A client
can request that $p_i$ execute a transaction $t \in T$. We consider what
happens when $t \in T_i$ and $t \notin T_i$ separately.

If $t \notin T_i$, then $p_i$ initiates a round of global coordination to
execute $t$. During global coordination, every server temporarily stops
processing and transitions to state $s = s_1 \join \ldots \join s_n$, the join
of every server's state. Then, every server speculatively executes $t$
transitioning to state $t(s)$. If $t(s)$ violates the invariant (i.e.  $\lnot
I(t(s))$), then every server aborts $t$ and reverts to state $s$.  Then, $p_i$
replies to the client. If $t(s)$ satisfies the invariant (i.e. $I(t(s))$), then
every server commits $t$ and remains in state $t(s)$. Every server then chooses
the first segment $(I_i, T_i) \in \Sigma$ such that $I_i(t(s))$ to be the new
active segment. Note that such a segment is guaranteed to exist because the
segment invariants cover $I$. Moreover, $\Sigma$ is ordered, so every server
will deterministically pick the same active segment. In fact, an invariant of
the system model is that at any given point of normal processing, every server
has the same active segment.

$t \in T_i$, then $p_i$ executes $t$ immediately and without coordination. If
$t(s_i)$ satisfies the \emph{active} invariant (i.e. $I_i(t(s_i))$), then $p_i$
commits $t$, stays in state $t(s_0)$, and replies to the client. If $t(s_i)$
violates the invariant (i.e.  $i.e. \lnot I(t(s_i))$), then $p_i$ aborts $t$,
reverts to state $s_i$, and replies to the client. If $t(s_i)$ satisfies the
invariant but violates the active invariant (i.e. $I(t(s_i))$ but $\lnot
I_i(t(s_i))$), then $p_i$ reverts to state $s_i$ and initiates a round of
global coordination to execute $t$, as described in the previous paragraph.
Transaction execution is summarized in \algoref{TxnExecution}.

\begin{algorithm}[t]
  \caption{Transaction execution in new system model}%
  \algolabel{TxnExecution}
  \begin{algorithmic}
    \If{$t \notin T_i$}
      \State Execute $t$ with global coordination
    \Else{}
      \If{$I_i(t(s_i))$} Commit $t$
      \ElsIf{$\lnot I(t(s_i))$} Abort $t$
      \Else{} Execute $t$ with global coordination
      \EndIf{}
    \EndIf{}
  \end{algorithmic}
\end{algorithm}

% why does this work

\subsection{Discussion}
% generalizes locking
% contiuous generalization of invariant-confluence

% The motivate the main idea behind segmented invariant-confluence, let's revisit
% \exampleref{Z2} but with a different start state.
%
% \begin{example}
%   Consider again the object $O = (\ints \times \ints, \join)$, transactions $T
%   = \set{t_{x+1}, t_{y-1}}$, and invariant $I = \setst{(x, y)}{xy \leq 0}$ from
%   \exampleref{Z2}, but now let our start state $s_0$ be $(-10, 10)$. Now, $O$
%   is \emph{not} \sTIconfluent{}. From $(-10, 10)$, we can reach the points $(0,
%   10)$ and $(10, 0)$, and merging these together we reach $(10, 10)$ which
%   violates the invariant.
%   %
%   However, all is not lost. If we're determined to make $O$
%   invariant-confluent, we can do so in one of two ways.
%   \begin{enumerate}
%     \item
%       We can modify the invariant. $O$ may not be \sTIconfluent, but it is
%       \sticonfluent{s_0}{T}{\setst{(x, y)}{x \leq 0, y \geq 0}}. If we eliminate
%       the fourth quadrant from the invariant, $O$ becomes invariant-confluent.
%     \item
%       We can disallow some transactions. $O$ may not be \sTIconfluent, but
%       it is \sticonfluent{s_0}{\set{t_{x+1}}}{I}. If we disallow the $t_{y-1}$
%       transaction, $O$ becomes invariant-confluent.
%   \end{enumerate}
%   While these two modifications are possible, neither is desirable. Artificially making some invariant-satisfying states unreachable---either by restricting the invariant or disallowing some transactions---is most likely a bit of a hack. Plus,
% \end{example}
